<!-- Set working directory -->
# Strings, Factors and Dataframe Handling

## What we will see today

Today we will look at some patient data, and we will answer these questions:

* Are data consistent?

* Which is the mean and sd values of glucose in each treatment group? How many patients we have per treatment group?

* Which group has the highest mean creatinine? (2 different datasets)


## Strings and Factors


Today we will work with a clinical dataset containing information about patients enrolled in a multi-site trial. Let's import it and, as always, **inspect it first**:

```{r}
library(tidyverse)

clinical_data <- read.csv("data/clinical_data.csv")
```

```{r}
str(clinical_data)
```

```{r}
summary(clinical_data)
```

```{r}
head(clinical_data)
```

You will notice that several columns contain text: `patient_id`, `sex`, `treatment`, `stage`, `smoking`, `site`, `diagnosis_code`. In R, text data can be represented in two fundamentally different ways: as **strings** (also called *characters*) or as **factors**. Let's understand the difference.

---

A **string** (or character) is simply a sequence of text. R treats it as a raw piece of text and does not impose any structure on it.

A **factor** is a *categorical variable*: it has a fixed, predefined set of allowed values called **levels**. Under the hood, R stores factors as integers (one integer per level) and keeps a table that maps integers to labels. This makes factors memory-efficient and very useful for statistical modelling (e.g., in regression, factors are automatically handled as categorical predictors) and for plotting purposes.

**When to use strings and when factors?**

| Use **string** when... | Use **factor** when... |
|---|---|
| Text is free-form or unpredictable | The variable takes a limited set of known values |
| You need to manipulate the text | Order matters (e.g., stage I < II < III < IV) |
| Unique identifiers (patient IDs) | Input to statistical models |

As some of our columns fall in the conditions for factors, let's convert them with the function `factor`:

```{r}
clinical_data$sex       <- factor(clinical_data$sex)
clinical_data$treatment <- factor(clinical_data$treatment)
clinical_data$site      <- factor(clinical_data$site)
clinical_data$smoking   <- factor(clinical_data$smoking)

# For stage, the categories have a natural ORDER: we declare an ordered factor
clinical_data$stage <- factor(clinical_data$stage,
                               levels = c("I", "II", "III", "IV"),
                               ordered = TRUE)

str(clinical_data)
```

Now `summary` gives us much more useful information on those columns:

```{r}
summary(clinical_data)
```

Notice how `summary` now shows the count per level for factors (e.g. how many Male vs Female), while before it just said `character`.

Moreover, we can see that in smoking column, we have different labels with the same meaning (never, Never, never smoker for never). R (as well as any programming language) is case sensitive, so it treats those 3 as different labels, even if for us they mean the same thing.

**IMPORTANT**: labels should be consistent across datasets when you fill them.

As we want to edit those to be consistent, we need to change the column back to be a character, with `as.character` function:

```{r}
clinical_data$smoking   <- as.character(clinical_data$smoking)
str(clinical_data)
```

We can now proceed.

### Operations on strings

The main package for string operations in R (and the one we will use throughout the course) is `stringr`, which is part of the `tidyverse`. All its functions start with `str_` to make them easy to remember.

#### Character substitution
Let's start by looking at how to change characters in a string: from "Curent" to "current" using `str_replace(vector, old_pattern, new_pattern)`.

```{r}
print(unique(clinical_data$smoking))
clinical_data$smoking <- str_replace(clinical_data$smoking, "Curent", "current")
print(unique(clinical_data$smoking))
```

Then we can unify "FORMER" and "former" by transforming the former to lowercase with `str_to_lower`:

```{r}
print(unique(clinical_data$smoking))
clinical_data$smoking <- str_to_lower(clinical_data$smoking)
print(unique(clinical_data$smoking))
```

We have converted both FORMER to former and Never to never. We lastly have to change "never smoker" to "never".

::: {.exercise #factor}

Change "never smoker" to "smoker" and make smoker as factor.

:::

<details>
  <summary>Solution</summary>

```{r}
clinical_data$smoking <- str_replace(clinical_data$smoking, "never smoker", "never")
clinical_data$smoking <- factor(clinical_data$smoking, levels = c("never", "former", "current"))

summary(clinical_data)
```

</details>

For sake of completeness, we could have done all of this with a single function `replace_values` (for new versions of R):

```{r, eval=FALSE}
clinical_data$smoking <- replace_values(as.character(clinical_data$smoking), c("never smoker", "Never") ~ "never",
                                        c("Curent") ~ "current",
                                        c("FORMER") ~ "former")

clinical_data$smoking = factor(clinical_data$smoking)
```

If it doesn't work, you could use:
```{r}
clinical_data = mutate(smoking = case_when(
    smoking %in% c("never smoker", "Never") ~ "never",
    smoking %in% c("FORMER", "Former", "Fomer") ~ "former",
    smoking == "Curent" ~ "current",
    .default = smoking),
    .data = clinical_data
    )

clinical_data$smoking = factor(clinical_data$smoking)
```


Q: *Why there are 2 Turin?*
A: Let's check in details the possible labels of that column with function `levels`:

```{r}
print(levels(clinical_data$site))
```

Here's the hitch, some have a trailing space after Turin. We can get rid of them using `str_trim` function:

```{r}
print(levels(clinical_data$site))
clinical_data$site <- factor(str_trim(clinical_data$site))

print(levels(clinical_data$site))
print(summary(clinical_data))
```

#### Other useful functions to handle strings

A few more `stringr` functions you will find very useful:

| Function | What it does |
|---|---|
| `str_detect(x, pattern)` | Returns TRUE/FALSE: does the string contain the pattern? |
| `str_starts(x, pattern)` | Does the string *start* with the pattern? |
| `str_ends(x, pattern)` | Does the string *end* with the pattern? |
| `str_pad(x, width, side, pad)` | Pads a string to a given width |
| `str_sub(x, start, end)` | Extracts a substring by position |

---

### Operations on multiple columns

#### Split a column and merge multiple columns

Sometimes a single column contains multiple pieces of information that could be split into separate columns. Look at `visit_date`:

```{r}
head(clinical_data$visit_date)
```

We can see that the year of visit varies a lot, maybe we should take it into account. We can split this into 3 separate columns using `separate` function:

```{r}
clinical_data <- separate(clinical_data,
                           col = visit_date,         # column to split
                           into = c("visit_year", "visit_month", "visit_day"), # new column names
                           sep = "-",                # separator
                           remove = FALSE)           # keep the original column

head(clinical_data)
```


The opposite operation — merging multiple columns into one — is done with `unite`:

```{r}
# example: unite back together
clinical_data <- unite(clinical_data,
                       col = "visit_year_month",         # name of the new column
                       visit_year, visit_month,           # columns to merge
                       sep = "-",                      # separator
                       remove = FALSE)                 # keep original columns

head(clinical_data)
```

---

### Column and row names

Every dataframe has **column names** (the variable names) and **row names** (by default a number, but we can change it).

To retrieve column names:

```{r}
colnames(clinical_data)
```

To rename specific columns, we can assign a new vector of names to `colnames`, or use the tidyverse function `rename` (we will see this with `dplyr` below):

```{r}
clinical_data = rename(clinical_data, "diastolic_bp_mmhg" = "dbp_mmhg", "systolic_bp_mmhg" = "sbp_mmhg")

colnames(clinical_data)
```

For row names:

```{r}
# by default, row names are row numbers
head(rownames(clinical_data))

# we could set patient_id as row names
rownames(clinical_data) <- clinical_data$patient_id

head(rownames(clinical_data))
```

However, **be careful** with row names: they are not a proper column of the dataframe, so they don't follow along easily in many tidyverse operations. It is generally better to keep your identifiers as a regular column.

---

## Dataframe Handling

The main verbs we will learn today are:

* `mutate` — add or modify columns
* `select` — choose which columns to keep
* `filter`/`filter_out` — keep/remove rows that satisfy a condition
* `group_by` + `summarize` / `reframe` — aggregate data by groups
* `left_join` / `inner_join` — merge two dataframes



### Create new columns with mutate

`mutate` creates new columns or modifies existing ones.

A key feature of `dplyr` is the **pipe operator** `|>` (or `%>%` in older tidyverse style). It passes the result of one operation directly to the next, making code much more readable:

You can read `|>` as *"and then"*.

```{r}
# Add a BMI column (we already know this formula!)
clinical_data <- clinical_data |> # take clinical_data and then
  mutate(height_m = height_cm / 100,
         BMI = weight_kg / height_m^2,
         BMI = round(BMI, 1))

head(clinical_data)
```

`mutate` can also create columns based on conditions, using `if_else` or `case_when`:

```{r}
# Classify BMI into categories
clinical_data <- clinical_data |>
  mutate(BMI_class = case_when(
    BMI < 18.5 ~ "Underweight",
    BMI < 25   ~ "Normal",
    BMI < 30   ~ "Overweight",
    BMI >= 30  ~ "Obese"
  ),
  BMI_class = factor(BMI_class, labels = c("Underweight", "Normal", "Overweight", "Obese"), ordered = T)
  )

summary(clinical_data$BMI_class)
```


---

### select

`select` chooses which **columns** to keep (or drop). This is useful when working with wide dataframes where you only need a subset of variables.

```{r}
# keep only demographic and clinical variables
demo_cols <- clinical_data |>
  dplyr::select(patient_id, site, age, sex, treatment, stage, BMI, BMI_class)

head(demo_cols)
```

To **drop** specific columns, use `-` before the column name:

```{r}

colnames(clinical_data)
# drop the helper/intermediate columns we created
clinical_data <- clinical_data |>
  dplyr::select(-visit_year_month)

colnames(clinical_data)
```


`select` also supports handy helper functions:

```{r}
# select all columns that start with "diag"
clinical_data |> dplyr::select(starts_with("visit")) |> head()

# select all columns that contain "bp"
clinical_data |> dplyr::select(contains("bp")) |> head()
```

---

### Boolean Operators

A **boolean** (or logical) is a data type that can only take two values: `TRUE` or `FALSE`. It is the backbone of filtering and conditional logic in R.

**A boolean is the answer to a yes/no question.**

For example: *"Is this patient in stage III or IV?"* → `TRUE` or `FALSE`.


#### Comparison operators

These operators compare two values and return a boolean:

| Operator | Meaning | Example | Result |
|---|---|---|---|
| `==` | equal to | `3 == 3` | `TRUE` |
| `!=` | not equal to | `3 != 4` | `TRUE` |
| `>`  | greater than | `5 > 3` | `TRUE` |
| `<`  | less than | `2 < 1` | `FALSE` |
| `>=` | greater or equal | `3 >= 3` | `TRUE` |
| `<=` | less or equal | `2 <= 1` | `FALSE` |

```{r}
# simple examples
print(10 > 5)
print(10 == 10)
print(10 != 10)
print(3.14 >= 3.14)
```

When applied to a **vector**, the comparison is done **element-wise** and returns a boolean vector:

```{r}
ages <- clinical_data$age
head(ages)

is_over_60 <- ages > 60
head(is_over_60)
```

Under the hood, TRUE is stored as a 1, while FALSE as 0. This is useful when we want to get the number of TRUE values of a vector. How? By summing up all the values:

```{r}
sum(is_over_60)
```


#### Logical operators

We often need to combine multiple conditions. This is done with **logical operators**:

| Operator | Meaning | Example |
|---|---|---|
| `&`  | AND — both conditions must be true | `age > 50 & sex == "Female"` |
| `\|` | OR — at least one condition must be true | `stage == "III" \| stage == "IV"` |
| `!`  | NOT — negates a condition | `!is.na(crp_ngml)` |

```{r}
# patients who are Female AND over 60
female_over60 <- (clinical_data$sex == "Female") & (clinical_data$age > 60)
sum(female_over60)
```

```{r}
# patients in advanced stage (III or IV)
advanced_stage <- (clinical_data$stage == "III") | (clinical_data$stage == "IV")
sum(advanced_stage)
```

```{r}
# NOT: patients who are NOT in the Control group
not_control <- clinical_data$treatment != "Control"
sum(not_control)

# equivalent using !
not_control_2 <- !(clinical_data$treatment == "Control")
sum(not_control_2)
```

A very useful operator when checking membership in a list of values is `%in%`:

```{r}
# patients from Milan or Rome (equivalent to == "Milan" | == "Rome")
big_sites <- clinical_data$site %in% c("Milan", "Rome")
sum(big_sites)
```

::: .{exercise #bool-sum}

How many patients have been diagnosed as stage I, in Turin in 2018?

:::

<details>
  <summary>Solution</summary>

```{r}
stageI_turin_2018 <- (clinical_data$site == "Turin") & (clinical_data$stage == "I") & (clinical_data$visit_year == 2018)
sum(stageI_turin_2018)
```
  
</details>

### filter and filter out

`filter` keeps only the **rows** that satisfy one or more conditions:

```{r}
# Keep only patients treated with Drug_A
drug_a_patients <- clinical_data |>
  filter(treatment == "Drug_A")

nrow(drug_a_patients)
head(drug_a_patients)
```

```{r}
# Patients in advanced stage AND not in control group
advanced_treated <- clinical_data |>
  filter((stage == "III" | stage == "IV") & treatment != "Control")

nrow(advanced_treated)
```

```{r}
# Using %in% for multiple values
milan_or_rome <- clinical_data |>
  filter(site %in% c("Milan", "Rome"))

nrow(milan_or_rome)
```

To **filter out** (i.e., *exclude* rows matching a condition), use `filter_out` (for new versions of R and tidyverse):

```{r}
# Exclude patients from Naples
no_milan <- clinical_data |>
  filter_out(site %in% c("Milan"))

nrow(no_milan)
```

If it doesn't work, you could use:

```{r}
no_milan = clinical_data %>% 
    filter(!(site == "Milan"))

nrow(no_milan)
```


Pipes shine when you chain multiple operations:

```{r}
# Female patients, over 55, from Milan, keep only key columns
subset_females <- clinical_data |>
  filter(sex == "Female", age > 55, site == "Milan") |>
  dplyr::select(patient_id, age, stage, treatment, BMI, glucose_mgdl)

head(subset_females)
nrow(subset_females)
```

Note: inside `filter`, multiple conditions separated by commas are treated as AND.

::: {.exercise #filter-exercise}

From the full dataset, extract patients who:

- are in stage II or III,
- have a glucose level between 90 and 130 mg/dL (inclusive),
- are assigned to Drug_A or Drug_B.

Keep only `patient_id`, `site`, `stage`, `treatment`, and `glucose_mgdl`.

:::

<details>
  <summary>Solution</summary>

```{r}
exercise_result <- clinical_data |>
  filter(stage %in% c("II", "III"),
         glucose_mgdl >= 90,
         glucose_mgdl <= 130,
         treatment %in% c("Drug_A", "Drug_B")) |>
  dplyr::select(patient_id, site, stage, treatment, glucose_mgdl)

head(exercise_result)
nrow(exercise_result)
```

</details>


---

### group_by, summarize and reframe

One of the most powerful operations in data analysis is **grouping** data and computing summaries per group. In R, this is done with `group_by` + `summarize`.

**How it works:**

1. `group_by(column)` tells R: *"split the dataframe into groups based on the values of this column"*
2. `summarize(...)` computes one summary value per group

<p align="center">
```
Original data:                 After group_by(treatment):

patient  treatment  glucose       GROUP: Control
PT-001   Control    105           PT-001  Control  105
PT-002   Drug_A     118    --->   PT-007  Control  98
PT-003   Drug_B     97            ...
PT-007   Control    98     --->   GROUP: Drug_A
...                               PT-002  Drug_A  118
                                  ...
```
</p>

We can perform several operation with this combo, and answer many questions. For example "*Which is the mean and sd values of glucose in each treatment group? How many patients we have per treatment group?*".

```{r}
# Mean glucose per treatment group
glucose_by_treatment <- clinical_data |>
  group_by(treatment) |>
  summarize(mean_glucose = mean(glucose_mgdl),
            sd_glucose   = sd(glucose_mgdl),
            n_patients   = n())

glucose_by_treatment
```

You can also group by multiple columns at once. 

"*For each treatment/sex combination, which is the mean BMI and the median of the age?*"

```{r}
# Summary by treatment AND sex
summary_by_group <- clinical_data |>
  group_by(treatment, sex) |>
  summarize(mean_BMI    = round(mean(BMI), 2),
            median_age  = median(age),
            n           = n(),
            .groups = "drop")   # good practice: drop grouping after summarize

summary_by_group
```


`summarize` returns **one row per group**. If you need to return **multiple rows per group** (e.g., all quantiles), use `reframe` instead:

```{r}
# Get quartiles of CRP per treatment group
crp_quantiles <- clinical_data |>
  group_by(treatment) |>
  reframe(quantile_value = quantile(crp_ngml, probs = c(0.25, 0.5, 0.75)),
          quantile_name  = c("Q1", "median", "Q3"))

crp_quantiles
```

Another useful function: `count`, which is a shortcut for `group_by + summarize(n = n())`:

```{r}
# Count patients per site and stage
clinical_data |>
  count(site, stage, sort = TRUE)
```


::: {.exercise #summarize-exercise}

For each combination of `site` and `sex`, compute:
- the number of patients,
- the mean systolic blood pressure,
- the proportion of hypertensive patients (hint: hypertension is when systolic blood pression is > 140).

:::

<details>
  <summary>Solution</summary>

```{r}
bp_summary <- clinical_data |>
    mutate(hypertension = systolic_bp_mmhg > 140) |>
    group_by(site, sex) |>
    summarize(
        n_patients          = n(),
        mean_systolic_bp_mmhg    = round(mean(systolic_bp_mmhg), 1),
        prop_hypertensive   = round(mean(hypertension), 3),
        .groups = "drop"
    )

bp_summary
```

</details>

---

### Join / Merge

In real-world analyses, data about the same subjects often lives in **separate files**: demographic data here, lab results there, outcomes somewhere else. To combine them, we **join** them on a shared key column (usually a patient ID).

Let's load the separate lab results file:

```{r}
lab_data <- read.csv("data/lab_results.csv")
str(lab_data)
head(lab_data)
```

The key column is `patient_id`. Not all patients have lab results. Depending on how we want to handle this, we choose different types of join:

| Join type | What it does |
|---|---|
| `left_join(x, y)` | Keep **all rows from x**, add y columns where available (NA if no match) |
| `right_join(x, y)` | Keep all rows from y |
| `inner_join(x, y)` | Keep only rows that exist in **both** x and y |
| `full_join(x, y)` | Keep all rows from both, NA where no match |


Let's see all approaches:

```{r}
# left_join: keep all patients, add lab data where available
left_df <- clinical_data |>
  left_join(lab_data, by = "patient_id")

nrow(left_df) 

# patients without lab data will have NA in lab columns
sum(is.na(left_df$hemoglobin_gdl))
```


Function `is.na()` returns TRUE when a value is NA, and FALSE if a value is an acutal value. Here an example:

```{r}
v1 = c(1, NA, 4) # create a vector with one NA

is.na(v1) # returns a vector with TRUE in position 2, and FALSE elsewhere
```


```{r}
summary(left_df)
```

<!-- Point at NA's -->


```{r}
# inner_join: keep only the 90 patients who have both clinical and lab data
matched_only <- clinical_data |>
  inner_join(lab_data, by = "patient_id")

nrow(matched_only)
```


In most clinical research scenarios, `left_join` is the most common choice: you start with your full patient list and enrich it with additional data, accepting that some patients may lack certain measurements (NAs).

::: {.exercise #join-exercise}

1. Perform a `left_join` of `clinical_data` with `lab_data`.
2. Among patients who have lab data available, compute the mean hemoglobin and mean creatinine per treatment group.
3. Which group has the highest mean creatinine?

:::

<details>
  <summary>Solution</summary>

```{r}
# 1. Left join
joined_df <- clinical_data |>
  left_join(lab_data, by = "patient_id")

# 2. Summarize (filter out NA first)
lab_summary <- joined_df |>
  filter(!is.na(hemoglobin_gdl)) |> # filter for what is NOT NA in hemoglobin
  group_by(treatment) |>
  summarize(mean_hemoglobin   = round(mean(hemoglobin_gdl), 2),
            mean_creatinine   = round(mean(creatinine_mgdl), 3),
            n                 = n(),
            .groups = "drop")

lab_summary

# 3. Highest mean creatinine
lab_summary |> arrange(desc(mean_creatinine))
```

</details>

---

## Save your results

After all this work, let's save the enriched and cleaned dataset:

```{r}
# Save the left-joined dataset with all derived variables
final_data <- clinical_data |>
  left_join(lab_data, by = "patient_id")

write.csv(final_data,
          file = "output/clinical_data_enriched.csv",
          row.names = FALSE)
```

---

## Home exercise

Try to do the following on your own:

1. Load `clinical_data_enriched.csv` from the output folder.
2. Clean up: keep only the columns `patient_id`, `site`, `sex`, `age`, `treatment`, `stage`, `BMI`, `BMI_class`, `glucose_mgdl`, `cholesterol_mgdl`, `crp_ngml`, `hemoglobin_gdl`, `creatinine_mgdl`.
3. Restrict to patients who have complete data (no NAs) — hint: use function `drop_na()`, see [this](https://www.r-bloggers.com/2024/12/how-to-use-drop_na-to-drop-rows-with-missing-values-in-r-a-complete-guide/).
4. For each stage, compute: number of patients, mean age, mean BMI, and proportion of obese patients.
5. Save the summary table to `output/summary_by_stage.csv`.

```{r}
library(tidyverse)

# 1. Load the enriched dataset
clinical_data_enriched <- read.csv("output/clinical_data_enriched.csv")

# 2. Keep only the columns of interest
clean_data <- clinical_data_enriched |>
  dplyr::select(patient_id, site, sex, age, treatment, stage,
         BMI, BMI_class, glucose_mgdl, cholesterol_mgdl,
         crp_ngml, hemoglobin_gdl, creatinine_mgdl)

# 3. Restrict to patients with complete data (no NAs)
clean_data <- clean_data |>
  drop_na()

cat("Patients with complete data:", nrow(clean_data), "\n")

# 4. Summary by stage
summary_by_stage <- clean_data |>
  group_by(stage) |>
  summarize(
    n_patients       = n(),
    mean_age         = round(mean(age), 1),
    mean_BMI         = round(mean(BMI), 2),
    prop_obese       = round(mean(BMI_class == "Obese"), 3),
    .groups = "drop"
  )

print(summary_by_stage)

# 5. Save
write.csv(summary_by_stage,
          file = "output/summary_by_stage.csv",
          row.names = FALSE)
```

---

## Bonus: how the clinical data were created

If you want to understand how we generated the dataset for this lesson, [download the R script](scripts/generate_mock_data.R) Notice the use of `set.seed` to make results reproducible.
