# For loops and If/Else statements

```{r}
library(tidyverse)
```

So far, every time we needed to repeat an operation on a different subset of data, we either copy-pasted the code or used tidyverse functions like `group_by`. Both approaches have limits: copy-pasting is error-prone and does not scale, while `group_by` only works inside a tidy pipeline.

**For loops** and **if/else statements** are the building blocks of any programming language and will let you automate repetitive tasks in a much more flexible way.

---

## For loops

The idea behind a for loop is simple: *repeat a block of code for each element of a sequence*.

The syntax in R is:

```
for (variable in sequence) {
    # code to repeat
}
```

- `sequence` is any vector or list: `c(1, 2, 3)`, `1:10`, a vector of filenames, ...
- `variable` is a name you choose; at each iteration it takes the next value of the sequence
- the code inside the `{}` is executed once per element

### A first example

The simplest possible loop: print each number from 1 to 5.

```{r}
for (i in 1:5) {
    print(i)
}
```

`i` here is just a conventional name (short for *index*), but you can call it whatever you want.

### Looping over a character vector

You can loop over any vector, not just numbers. Here we loop over a vector of stop codons:

```{r}
stop_codons <- c("UAG", "UAA", "UGA")

for (stop_c in stop_codons) {
    print(paste("Stop codon in this iteration is:", stop_c))
    Sys.sleep(2)
}
```

Let's say we have a vector of area values in mm² and we want to convert each one to µm²:

```{r}
areas_mm2 <- c(0.0042, 0.0031, 0.0058, 0.0049, 0.0037)

for (area in areas_mm2) {
    area_um2 <- area * 1e6
    print(paste(area, "mm² =", area_um2, "µm²"))
}
```

### Collecting results: the importance of initialising a container

In the example above, we just printed the results. Most of the time you want to **store** them. The right way to do this is to create an empty vector *before* the loop and fill it at each iteration.

```{r}
areas_um2 <- c()   # empty container

for (i in 1:length(areas_mm2)) {
  areas_um2[i] <- areas_mm2[i] * 1e6
}

areas_um2
```

Notice that we are now looping over the *index* `i` (from 1 to the length of the vector) rather than the values directly. This is very useful when you need to refer to the position of an element, for example to store the result in the same position of another vector.


### Looping over a list of dataframes

In biology, a very common scenario is having multiple dataframes — one per sample, one per experiment, etc. — and wanting to apply the same operation to each of them. We can store them in a **list** and loop over the list.

Let's create a small example with 3 mini-dataframes:

```{r}
df1 <- data.frame(group = c("wt", "KO"), intensity = c(450, 1420), area = c(0.005, 0.0048))
df2 <- data.frame(group = c("wt", "KO"), intensity = c(510, 1380), area = c(0.0052, 0.0051))
df3 <- data.frame(group = c("wt", "KO"), intensity = c(490, 1600), area = c(0.0049, 0.0047))

sample_list <- list(df1, df2, df3)
```

Now we loop over the list and calculate the normalised intensity for each dataframe:

```{r}
for (i in 1:length(sample_list)) {
  
    df <- sample_list[[i]]   # [[]] extracts one element from a list
    df$norm_intensity <- df$intensity / (df$area * 1e6)
    sample_list[[i]] <- df   # store the modified df back in the list
  
}

# Let's check the first element
sample_list[[1]]
```

Note the `[[]]` double brackets: to extract a single element from a list you always use `[[]]`, not `[]`.

### Naming the elements of a list

It is good practice to name the elements of a list, so that you can refer to them by name later:

```{r}
names(sample_list) <- c("sample_1", "sample_2", "sample_3")

sample_list[["sample_1"]]
```

---

## If/Else statements

An `if/else` statement lets you execute different code depending on whether a condition is `TRUE` or `FALSE`.

The syntax is:

```
if (condition) {
    # code if condition is TRUE
} else {
    # code if condition is FALSE
}
```

### A simple example

```{r}
x <- 15

if (x > 10) {
    print("x is greater than 10")
} else {
    print("x is 10 or less")
}
```

Change `x` to `5` and re-run to see the other branch.

### else if: more than two cases

When you have more than two possible outcomes, you can chain conditions with `else if`:

```{r}
p_value <- 0.03

if (p_value < 0.001) {
    print("Highly significant (***)")
} else if (p_value < 0.01) {
    print("Very significant (**)")
} else if (p_value < 0.05) {
    print("Significant (*)")
} else {
    print("Not significant")
}
```

R evaluates the conditions in order and executes the **first** block whose condition is `TRUE`. Once a match is found, the rest are skipped.

## Putting it all together: reading files in a loop

In the next session we will do this for real with our immunofluorescence data, but let's preview the full pattern here. The typical workflow is:

1. Get the list of files to process
2. Create an empty list to collect results
3. Loop over the files: read each one, process it, store it in the list
4. Combine all dataframes into one big dataframe with `bind_rows`

```{r}
# 1. Get the list of files
files <- list.files(path = "data/day-5", pattern = "*.csv", full.names = TRUE)
files
```

```{r}
# 2. Empty list to collect results
results_list <- list()

# 3. Loop
for (i in 1:length(files)) {
  
  # Read the file
  df <- read.csv(files[i])
  
  # Add a column with the sample name (useful to know where each row comes from)
  df$sample <- basename(files[i])
  
  # Convert area from mm2 to um2
  df$area_um2 <- df$area * 1e6
  
  # Calculate normalised intensity
  df$norm_intensity <- df$intensity / df$area_um2
  
  # Store in the list
  results_list[[i]] <- df
  
}

# 4. Combine into one dataframe
results_df <- bind_rows(results_list)

head(results_df)
```

We now have a single tidy dataframe with all samples, ready for statistical analysis.

```{r}
# Quick sanity check: how many rows per sample?
results_df %>%
  count(sample)
```

```{r}
# And a first look at the normalised intensity by group
ggplot(results_df) +
  geom_boxplot(aes(x = group, y = norm_intensity, fill = group)) +
  labs(title = "Normalised intensity by group",
       y = "Normalised intensity (a.u. / µm²)",
       x = "Group") +
  theme_classic()
```

