---
title: "Strings, Factors and Dataframe Handling"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Strings, Factors and Dataframe Handling

## What we will see today

Today we will look at some patient data, and we will answer these questions:

* Are data consistent?

* Which is the mean and sd values of glucose in each treatment group? How many patients we have per treatment group?

* Which group has the highest mean creatinine? (2 different datasets)


## Strings and Factors


Today we will work with a clinical dataset containing information about patients enrolled in a multi-site trial. Let's import it and, as always, **inspect it first**:

```{r}
library(tidyverse)

```

```{r}

```

```{r}

```

```{r}

```

You will notice that several columns contain text: `patient_id`, `sex`, `treatment`, `stage`, `smoking`, `site`, `diagnosis_code`. In R, text data can be represented in two fundamentally different ways: as **strings** (also called *characters*) or as **factors**. Let's understand the difference.

---

A **string** (or character) is simply a sequence of text. R treats it as a raw piece of text and does not impose any structure on it.

A **factor** is a *categorical variable*: it has a fixed, predefined set of allowed values called **levels**. Under the hood, R stores factors as integers (one integer per level) and keeps a table that maps integers to labels. This makes factors memory-efficient and very useful for statistical modelling (e.g., in regression, factors are automatically handled as categorical predictors) and for plotting purposes.

**When to use strings and when factors?**

| Use **string** when... | Use **factor** when... |
|---|---|
| Text is free-form or unpredictable | The variable takes a limited set of known values |
| You need to manipulate the text | Order matters (e.g., stage I < II < III < IV) |
| Unique identifiers (patient IDs) | Input to statistical models |

As some of our columns fall in the conditions for factors, let's convert them with the function `factor`:

```{r}

```

Now `summary` gives us much more useful information on those columns:

```{r}

```

Notice how `summary` now shows the count per level for factors (e.g. how many Male vs Female), while before it just said `character`.

Moreover, we can see that in smoking column, we have different labels with the same meaning (never, Never, never smoker for never). R (as well as any programming language) is case sensitive, so it treats those 3 as different labels, even if for us they mean the same thing.

**IMPORTANT**: labels should be consistent across datasets when you fill them.

As we want to edit those to be consistent, we need to change the column back to be a character, with `as.character` function:

```{r}

```

We can now proceed.

### Operations on strings

The main package for string operations in R (and the one we will use throughout the course) is `stringr`, which is part of the `tidyverse`. All its functions start with `str_` to make them easy to remember.

#### Character substitution
Let's start by looking at how to change characters in a string: from "Curent" to "current" using `str_replace(vector, old_pattern, new_pattern)`.

```{r}

```

Then we can unify "FORMER" and "former" by transforming the former to lowercase with `str_to_lower`:

```{r}

```

We have converted both FORMER to former and Never to never. We lastly have to change "never smoker" to "smoker".

::: {.exercise #factor}

Change "never smoker" to "smoker" and make smoker as factor.

:::

<details>
  <summary>Solution</summary>

```{r}

```

</details>

For sake of completeness, we could have done all of this with a single function `replace_values`:

```{r, eval=FALSE}

```


Q: *Why there are 2 Turin?*
A: Let's check in details the possible labels of that column with function `levels`:

```{r}

```

Here's the hitch, some have a trailing space after Turin. We can get rid of them using `str_trim` function:

```{r}

```

#### Other useful functions to handle strings

A few more `stringr` functions you will find very useful:

| Function | What it does |
|---|---|
| `str_detect(x, pattern)` | Returns TRUE/FALSE: does the string contain the pattern? |
| `str_starts(x, pattern)` | Does the string *start* with the pattern? |
| `str_ends(x, pattern)` | Does the string *end* with the pattern? |
| `str_pad(x, width, side, pad)` | Pads a string to a given width |
| `str_sub(x, start, end)` | Extracts a substring by position |

---

### Operations on multiple columns

#### Split a column and merge multiple columns

Sometimes a single column contains multiple pieces of information that could be split into separate columns. Look at `visit_date`:

```{r}

```

We can see that the year of visit varies a lot, maybe we should take it into account. We can split this into 3 separate columns using `separate` function:

```{r}

```


The opposite operation — merging multiple columns into one — is done with `unite`:

```{r}

```

---

### Column and row names

Every dataframe has **column names** (the variable names) and **row names** (by default a number, but we can change it).

To retrieve column names:

```{r}

```

To rename specific columns, we can assign a new vector of names to `colnames`, or use the tidyverse function `rename` (we will see this with `dplyr` below):

```{r}

```

For row names:

```{r}

```

However, **be careful** with row names: they are not a proper column of the dataframe, so they don't follow along easily in many tidyverse operations. It is generally better to keep your identifiers as a regular column.

---

## Dataframe Handling

The main verbs we will learn today are:

* `mutate` — add or modify columns
* `select` — choose which columns to keep
* `filter`/`filter_out` — keep/remove rows that satisfy a condition
* `group_by` + `summarize` / `reframe` — aggregate data by groups
* `left_join` / `inner_join` — merge two dataframes



### Create new columns with mutate

`mutate` creates new columns or modifies existing ones.

A key feature of `dplyr` is the **pipe operator** `|>` (or `%>%` in older tidyverse style). It passes the result of one operation directly to the next, making code much more readable:

You can read `|>` as *"and then"*.

```{r}

```

`mutate` can also create columns based on conditions, using `if_else` or `case_when`:

```{r}

```


---

### select

`select` chooses which **columns** to keep (or drop). This is useful when working with wide dataframes where you only need a subset of variables.

```{r}

```

To **drop** specific columns, use `-` before the column name:

```{r}

```


---

### Boolean Operators

A **boolean** (or logical) is a data type that can only take two values: `TRUE` or `FALSE`. It is the backbone of filtering and conditional logic in R.

**A boolean is the answer to a yes/no question.**

For example: *"Is this patient in stage III or IV?"* → `TRUE` or `FALSE`.


#### Comparison operators

These operators compare two values and return a boolean:

| Operator | Meaning | Example | Result |
|---|---|---|---|
| `==` | equal to | `3 == 3` | `TRUE` |
| `!=` | not equal to | `3 != 4` | `TRUE` |
| `>`  | greater than | `5 > 3` | `TRUE` |
| `<`  | less than | `2 < 1` | `FALSE` |
| `>=` | greater or equal | `3 >= 3` | `TRUE` |
| `<=` | less or equal | `2 <= 1` | `FALSE` |

```{r}
# simple examples

```

When applied to a **vector**, the comparison is done **element-wise** and returns a boolean vector:

```{r}

```

Under the hood, TRUE is stored as a 1, while FALSE as 0. This is useful when we want to get the number of TRUE values of a vector. How? By summing up all the values:

```{r}

```


#### Logical operators

We often need to combine multiple conditions. This is done with **logical operators**:

| Operator | Meaning | Example |
|---|---|---|
| `&`  | AND — both conditions must be true | `age > 50 & sex == "Female"` |
| `\|` | OR — at least one condition must be true | `stage == "III" \| stage == "IV"` |
| `!`  | NOT — negates a condition | `!is.na(crp_ngml)` |

```{r}

```

```{r}

```

```{r}

```

A very useful operator when checking membership in a list of values is `%in%`:

```{r}

```


**Exercise**
How many patients have been diagnosed as stage I, in Turin in 2018?


```{r}

```
  

### filter and filter out

`filter` keeps only the **rows** that satisfy one or more conditions:

```{r}

```

```{r}

```

```{r}

```

To **filter out** (i.e., *exclude* rows matching a condition), use `filter_out`:

```{r}

```

Pipes shine when you chain multiple operations:

```{r}

```

Note: inside `filter`, multiple conditions separated by commas are treated as AND.

**Exercise**

From the full dataset, extract patients who:

- are in stage II or III,
- have a glucose level between 90 and 130 mg/dL (inclusive),
- are assigned to Drug_A or Drug_B.

Keep only `patient_id`, `site`, `stage`, `treatment`, and `glucose_mgdl`.


```{r}

```



---

### group_by, summarize and reframe

One of the most powerful operations in data analysis is **grouping** data and computing summaries per group. In R, this is done with `group_by` + `summarize`.

**How it works:**

1. `group_by(column)` tells R: *"split the dataframe into groups based on the values of this column"*
2. `summarize(...)` computes one summary value per group

<p align="center">
```
Original data:                 After group_by(treatment):

patient  treatment  glucose       GROUP: Control
PT-001   Control    105           PT-001  Control  105
PT-002   Drug_A     118    --->   PT-007  Control  98
PT-003   Drug_B     97            ...
PT-007   Control    98     --->   GROUP: Drug_A
...                               PT-002  Drug_A  118
                                  ...
```
</p>

We can perform several operation with this combo, and answer many questions. For example "*Which is the mean and sd values of glucose in each treatment group? How many patients we have per treatment group?*".

```{r}

```

You can also group by multiple columns at once. 

"*For each treatment/sex combination, which is the mean BMI and the median of the age?*"

```{r}

```



**Exercise**

For each combination of `site` and `sex`, compute:
- the number of patients,
- the mean systolic blood pressure,
- the proportion of hypertensive patients (hint: hypertension is when systolic blood pression is > 140).


```{r}

```



---

### Join / Merge

In real-world analyses, data about the same subjects often lives in **separate files**: demographic data here, lab results there, outcomes somewhere else. To combine them, we **join** them on a shared key column (usually a patient ID).

Let's load the separate lab results file:

```{r}

```

The key column is `patient_id`. Not all patients have lab results. Depending on how we want to handle this, we choose different types of join:

| Join type | What it does |
|---|---|
| `left_join(x, y)` | Keep **all rows from x**, add y columns where available (NA if no match) |
| `right_join(x, y)` | Keep all rows from y |
| `inner_join(x, y)` | Keep only rows that exist in **both** x and y |
| `full_join(x, y)` | Keep all rows from both, NA where no match |


Let's see all approaches:

```{r}

```

```{r}

```



```{r}

```


In most clinical research scenarios, `left_join` is the most common choice: you start with your full patient list and enrich it with additional data, accepting that some patients may lack certain measurements (NAs).

**Exercise**
1. Perform a `left_join` of `clinical_data` with `lab_data`.
2. Among patients who have lab data available, compute the mean hemoglobin and mean creatinine per treatment group.
3. Which group has the highest mean creatinine?


```{r}

```


---

## Save results

After all this work, let's save the enriched and cleaned dataset:

```{r}

```

---

## Home exercise

Try to do the following on your own:

1. Load `clinical_data_enriched.csv` from the output folder.
2. Clean up: keep only the columns `patient_id`, `site`, `sex`, `age`, `treatment`, `stage`, `BMI`, `BMI_class`, `glucose_mgdl`, `cholesterol_mgdl`, `crp_ngml`, `hemoglobin_gdl`, `creatinine_mgdl`.
3. Restrict to patients who have complete data (no NAs) — hint: use function `drop_na()`, see [this](https://www.r-bloggers.com/2024/12/how-to-use-drop_na-to-drop-rows-with-missing-values-in-r-a-complete-guide/).
4. For each stage, compute: number of patients, mean age, mean BMI, and proportion of obese patients.
5. Save the summary table to `output/summary_by_stage.csv`.


